---
title: "Code for MTHM601 report"
output:
  pdf_document: default
  html_document: default
date: "2025-01-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(patchwork)
library(dplyr)
library(lubridate)
library(ggpubr)
library(mgcv)
library(sf)
library(tmap)
library(leaflet)
library(spatstat)
library(flextable)
library(MASS)
library(lme4)
library(emmeans)
```

## Load Roadkill dataset

```{r}
roadkill_full <- read.csv("records-2024-12-03.csv")
```

## Data Wrangling

### Rename columns

Using the dplyr package, rename columns so they are more manageable.

```{r}
head(roadkill_full) #view column names

rk <- roadkill_full %>% #rename complex names to more manageable ones to reduce errors
  rename(
    species = Common.name,
    date = Start.date,
    day = Start.date.day,
    month = Start.date.month,
    year = Start.date.year,
    lat = Latitude..WGS84.,
    long = Longitude..WGS84.,
    country = State.Province
  )

head(rk) #view column names for new dataset
```

Create dataset with only necessary columns. Using the dplyr package.

```{r}
roadkill <- rk %>% dplyr::select(species, Taxon.Rank, date, day, month, year, lat, long, country, Kingdom, Phylum, Class, Order, Family, Genus) #only keep relevant columns


```

### Add a season column

I am interested in seasonal variation, not just the exact date so I am going to use an if-else function to create a season column.

```{r}
get_season <- function(month) { #if-else function to turn months into their corresponding season
  if (month %in% c(12, 1, 2)) {
    return("Winter")
  } else if (month %in% c(3, 4, 5)) {
    return("Spring")
  } else if (month %in% c(6, 7, 8)) {
    return("Summer")
  } else if (month %in% c(9, 10, 11)) {
    return("Autumn")
  } else {
    return(NA)
  }
}


roadkill <- roadkill %>%
  mutate(season = sapply(month, get_season)) #create new season column by applying the if-else function to the month column


head(roadkill)


```

Noticed that date is coming up as NA - so I will combine the day, month and year columns to create a new date column

```{r}
roadkill <- roadkill %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"), format = "%Y-%m-%d")) #change format of year to make sure it shows up correctly

head(roadkill)


```

### Format data types

```{r}
#format data type for each column specific to the contents of each column
roadkill$day <- as.integer(roadkill$day)
roadkill$month <- as.integer(roadkill$month)
roadkill$year <- as.integer(roadkill$year)
roadkill$species <- as.factor(roadkill$species)
roadkill$Taxon.Rank <-as.factor(roadkill$Taxon.Rank)
roadkill$lat <- as.numeric(roadkill$lat)
roadkill$long <- as.numeric(roadkill$long)
roadkill$country <-as.factor(roadkill$country)
roadkill$Kingdom <-as.factor(roadkill$Kingdom)
roadkill$Phylum <-as.factor(roadkill$Phylum)
roadkill$Class <-as.factor(roadkill$Class)
roadkill$Order <-as.factor(roadkill$Order)
roadkill$Family <-as.factor(roadkill$Family)
roadkill$Genus <-as.factor(roadkill$Genus)
roadkill$season <-as.factor(roadkill$season)

str(roadkill)
```

### Remove all birds and reptiles - only keep mammals

```{r}

mammal_roadkill <- roadkill %>%
  filter(grepl("mammalia", Class, ignore.case = TRUE)) #Only keep mammal reports

head(mammal_roadkill)
str(mammal_roadkill)


```

### Remove any NAs or blanks

```{r}

mammal_roadkill <- mammal_roadkill %>%
  mutate(species = trimws(species)) %>% 
  filter(species != "", !is.na(species)) #remove any NAs and blanks from the dataset

head(mammal_roadkill)
str(mammal_roadkill)

```

### Unique species

```{r}
unique(mammal_roadkill$species) #list all unique values in the species column
```

### Remove "indet.deer" and "polecat-ferret hybrid" and "rabbit and hares"

Because they are not to a species level, I opted to remove any rows that were called these things.

```{r}
mammal_roadkill <- mammal_roadkill %>%
  filter(!species %in% c("Indet. Deer", "Polecat-Ferret", "rabbits and hares")) #remove these reports
```

### Standardise species names

Some species were hard to identify to species level, so I decided to group them. Also grey squirrel and squirrel, I grouped as Red Squirrels are rare in most parts of the country.

```{r}
#some are recorded with multiple names, combine them into generalised names
mammal_roadkill <- mammal_roadkill %>%
  mutate(species = case_when(
    species %in% c("Eastern Grey Squirrel", "Squirrel") ~ "Grey Squirrel",
    species %in% c("Brown Rat", "Rat spp.") ~ "Rats",
    species %in% c("Wood Mouse", "House Mouse", "Small Mouse", "Yellow-necked Mouse") ~ "Mice",
    species %in% c("Bat", "Pipistrelle Bat species", "Soprano Pipistrelle", "Brown Long-eared Bat", "Pipistrelle") ~ "Bats",
    species %in% c("Field Vole", "Bank Vole", "Vole") ~ "Voles",
    species %in% c("Brown Hare", "Mountain Hare", "Irish Hare") ~ "Hares",
    TRUE ~ species 
  ))


#view the updated dataset
head(mammal_roadkill)
unique(mammal_roadkill$species)

```


### Exploratory Data Visualisation

#### Monthly reports

```{r}
monthly_yearly_data <- mammal_roadkill %>% #group data by year and month and then count roadkill reports for that year-month combination
  group_by(year, month) %>%
  summarise(total_roadkill = n())

monthly_yearly_data <- monthly_yearly_data %>%
  mutate(month = factor(month, levels = 1:12, labels = month.name)) #convert month to factor value

print(monthly_yearly_data)


```

```{r}
#plot the trends over the year - colour = year
p1 <- ggplot(monthly_yearly_data, aes(x = month, y = total_roadkill, group = year, color = factor(year))) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Monthly Roadkill Trends by Year",
    x = "Month",
    y = "Total Roadkill",
    color = "Year"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1
```

```{r}
monthly_roadkill <- mammal_roadkill %>%
  group_by(year, month) %>%
  summarise(reports = n(), .groups = "drop") #group and summarise data

monthly_roadkill <- monthly_roadkill %>%
  mutate(date = as.Date(paste(year, month, 1, sep = "-"))) #add a date column for easy plotting

#plot trend in reports from 2014 to 2024 on one axis
p2 <- ggplot(monthly_roadkill, aes(x = date, y = reports)) +
  geom_line(color = "black", linewidth = 0.5) +
  theme_minimal() +
  labs(
    title = "(a)",
    x= NULL,
    y = NULL  ) +
  ylim(0,2500)+
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2
```

## Scale spike in 2019

Following the methods from Raymond et al. 2021, I scaled the data based on monthly proportions and then I subsampled it 10,000 times and took the mean.

```{r}
#scale the data of monthly totals based on percentages of annual totals
scaled_data <- mammal_roadkill %>%
  group_by(species, year, month) %>%
  summarise(monthly_total = n(), .groups = "drop") %>%  
  group_by(species, year) %>%
  mutate(
    annual_total = sum(monthly_total), 
    monthly_percentage = round((monthly_total / annual_total) * 100, 1) 
  ) %>%
  ungroup()

head(scaled_data)


```

```{r}

set.seed(222) #for reproducibility

adjusted_data <- scaled_data %>% #adjust monthly totals
  group_by(species) %>%
  mutate(adjusted_count = case_when(
    year == 2019 & month == 7 & monthly_total > 1 ~ round(monthly_total * 2 / 3), 
    year == 2019 & month == 8 & monthly_total > 1 ~ round(monthly_total / 2),    
    TRUE ~ monthly_total 
  )) %>%
  ungroup()

#Subsampling repeated 10,000 times and taking the mean count
adjusted_data <- adjusted_data %>%
  group_by(species, year, month) %>%
  summarise(
    mean_adjusted_count = mean(replicate(10000, {
      if (year == 2019 & month == 7 && monthly_total > 1) {
        round(monthly_total * 2 / 3)
      } else if (year == 2019 & month == 8 && monthly_total > 1) {
        round(monthly_total / 2)
      } else {
        monthly_total
      }
    })),
    .groups = "drop"
  )

head(adjusted_data)


```

Plot this adjusted data

```{r}
total_adjusted_data <- adjusted_data %>% #group and summarise ocunts and add date column 
  group_by(year, month) %>%
  summarise(total_reports = sum(mean_adjusted_count, na.rm = TRUE), .groups = "drop") %>%
  mutate(date = ymd(paste(year, month, "01", sep = "-")))

#plot the adjusted data
p3 <- ggplot(total_adjusted_data, aes(x = date, y = total_reports)) +
  geom_line(color = "black", linewidth = 0.5) +
  theme_minimal() +
  labs(
    title = "(b)",
    x = NULL,
    y = "Reports"
  ) +
  ylim(0,2500)+
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p3
```

I decided this still resulted in too big of a spike so I will remove July - September 2019 and then ttry imputing values.

### Remove July - September 2019

```{r}

mammal_roadkill_no_spike <- mammal_roadkill %>% #remove July and August 2019 
  filter(!(format(date, "%Y-%m") %in% c("2019-07", "2019-08", "2019-09")))

```

### Impute values for July - September 2019

I will turn it into monthly values and then add blank rows for July - September 2019 and then impute values.

```{r}

monthly_no_spike <- mammal_roadkill_no_spike %>% #summarising monthly
  group_by(year, month) %>%
  summarise(total_roadkill = n())

monthly_no_spike <- monthly_no_spike %>% #add month names
  mutate(month = factor(month, levels = 1:12, labels = month.name))

print(monthly_no_spike)

imputed_rk <- monthly_no_spike %>% #impute missing data
  mutate(month = factor(month, levels = month.name)) %>%
  group_by(month) %>%  
  mutate(monthly_mean = mean(total_roadkill, na.rm = TRUE),  
         value = ifelse(is.na(total_roadkill) & year == 2019, monthly_mean, total_roadkill)) %>%  
  dplyr::select(-monthly_mean) #delete temporary column

print(imputed_rk)
```

Change of plan, I realised this won't work for what I need as I don't just want monthly values. I am going to calculate the mean for all July's and August's and September's (excluding 2019) and from that mean, I will randomly select that amount of reports from July and August and September in 2019 and keep those, discarding the others. Because I'm not assessing anything at a species level, it doesn't matter if I lose individual species.

```{r}
july_mean <- mammal_roadkill_no_spike %>%
  filter(format(date, "%m") == "07") %>%  #filter for July data
  group_by(year = format(date, "%Y")) %>%  
  summarise(july_count = n()) %>% #counts per year
  summarise(mean_july_reports = mean(july_count, na.rm = TRUE)) #take the mean

print(round(july_mean))

#repeat the same for August and September
aug_mean <- mammal_roadkill_no_spike %>%
  filter(format(date, "%m") == "08") %>%  
  group_by(year = format(date, "%Y")) %>%  
  summarise(aug_count = n()) %>%  
  summarise(mean_aug_reports = mean(aug_count, na.rm = TRUE))  

print(round(aug_mean))

sept_mean <- mammal_roadkill_no_spike %>%
  filter(format(date, "%m") == "09") %>%  
  group_by(year = format(date, "%Y")) %>%  
  summarise(sept_count = n()) %>%  
  summarise(mean_sept_reports = mean(sept_count, na.rm = TRUE))  

print(round(sept_mean))
```

### Choose rows from 2019

```{r}
set.seed(2108)
random_sample_july_2019 <- mammal_roadkill %>%
  filter(format(date, "%Y-%m") == "2019-07") %>% #filter for July 2019
  sample_n(363) #randomly sample the number from the mean of Julys without July 2019

print(random_sample_july_2019)

#repeat for August and September
set.seed(2108)
random_sample_aug_2019 <- mammal_roadkill %>%
  filter(format(date, "%Y-%m") == "2019-08") %>%
  sample_n(383)

print(random_sample_aug_2019)

set.seed(2108)
random_sample_sept_2019 <- mammal_roadkill %>%
  filter(format(date, "%Y-%m") == "2019-09") %>%
  sample_n(420)

print(random_sample_sept_2019)
```

### Add these into the dataset

```{r}
mammal_no_spike <- mammal_roadkill_no_spike %>% #Add these to the dataset
  bind_rows(random_sample_july_2019, random_sample_aug_2019, random_sample_sept_2019) %>%
  arrange(date)


```

### Plot

```{r}
monthly <- mammal_no_spike %>%
  group_by(year, month) %>%
  summarise(reports = n(), .groups = "drop")

monthly <- monthly %>%
  mutate(date = as.Date(paste(year, month, 1, sep = "-")))


p4 <- ggplot(monthly, aes(x = date, y = reports)) +
  geom_line(color = "black", linewidth = 0.5) +
  theme_minimal() +
  labs(
    title = "(c)",
    x= NULL,
    y = NULL  ) +
  ylim(0, 2500) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4
```

Plot all 3 options together without their x labels.

```{r}
(p2 + theme(axis.title.x = element_blank(), axis.text.x = element_blank())) /
(p3 + theme(axis.title.x = element_blank(), axis.text.x = element_blank())) /
p4
```

## Met-Office data

Rainfall and temperature data from the met office don't download to excel files, just txt URLs so I have to load these datasets in a different way.

### Load dataset - temperature

Load dataset, turn it into a table, filter to the years I am interested in and then rearrange the table so that it is 'tidy' and each row reflects one month.

```{r}
tempURL <- "https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Tmean/date/UK.txt"

temp <- read.table(tempURL, header = TRUE, skip = 5, stringsAsFactors = FALSE) #turn data into a table and remove the first lines that are not part of the dataset

filtered_temp <- temp %>% #filter to only include 2014 - 2024
  filter(year >= 2014 & year <= 2024)

data_long <- filtered_temp %>% #mutate from wide-format to long format
  pivot_longer(
    cols = jan:dec,  # Month columns
    names_to = "month",
    values_to = "mean_temperature"
  ) %>%
  mutate(
    month = match(month, c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")),  
    year = as.numeric(year) 
  )

```

### Load dataset - rain

And now the same for the rain dataset.

```{r}
rainURL <- "https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Rainfall/date/UK.txt" 

rain <- read.table(rainURL, header = TRUE, skip = 5, stringsAsFactors = FALSE)

filtered_rain <- rain %>%
  filter(year >= 2014 & year <= 2024)


long_rain <- filtered_rain %>%
  pivot_longer(
    cols = jan:dec,  
    names_to = "month",
    values_to = "rainfall"
  ) %>%
  mutate(
    month = match(month, c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")),
    year = as.numeric(year)
  )
```

### Plot temperature

```{r}
t1 <- ggplot(data_long, aes(x = month, y = mean_temperature, group = year, color = factor(year))) +
  geom_smooth(se = FALSE, method = "loess", linewidth = 1) +
  geom_vline(xintercept = c(2.5, 5.5, 8.5, 11.5), linetype = "dashed", color = "black", linewidth = 0.8) + #lines to break up the seasons
  scale_x_continuous(
    breaks = 1:12,  #position for 1 to 12 for months
    labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  ) +
  labs(
    title = "Mean Temperature Over Months (2014-2024, UK)",
    x = "Month",
    y = "Mean Temperature (°C)",
    color = "Year"
  ) +
  theme_minimal()

t1

```

#### Plot rain

```{r}
t2 <- ggplot(long_rain, aes(x = month, y = rainfall, group = year, color = factor(year))) +
  geom_smooth(se = FALSE, method = "loess", linewidth = 1) +
  geom_vline(xintercept = c(2.5, 5.5, 8.5, 11.5), linetype = "dashed", color = "black", linewidth = 0.8) +
  scale_x_continuous(
    breaks = 1:12,  
    labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")  
  ) +
  labs(
    title = "Rainfall Over Months (2014-2024, UK)",
    x = "Month",
    y = "Rainfall (mm)",
    color = "Year"
  ) +
  theme_minimal()

t2
```

```{r}
t1/t2 + plot_layout(guides = "collect") #only one legend
```

## Combine roadkill and climate data

```{r}
climate_data <- long_rain %>% #combine rainfall data and roadkill
  full_join(data_long, by = c("year", "month"))


roadkill_climate <- mammal_no_spike %>% #add climate data 
  left_join(climate_data, by = c("year", "month"))

roadkill_climate <- roadkill_climate %>%
  mutate(across(ends_with(".x"), as.numeric)) %>%   
  mutate(across(ends_with(".y"), as.numeric)) 


roadkill_with_climate <- roadkill_climate %>%
  mutate(
    seasonal_rainfall = case_when(
      season == "Winter" ~ win.x + win.y,
      season == "Spring" ~ spr.x + spr.y,
      season == "Summer" ~ sum.x + sum.y,
      season == "Autumn" ~ aut.x + aut.y,
      TRUE ~ NA_real_  # Default for unexpected cases
    )
  ) %>%
  dplyr::select(-win.x, -win.y, -spr.x, -spr.y, -sum.x, -sum.y, -aut.x, -aut.y)


roadkill_counts <- roadkill_with_climate %>% #summarise counts 
  group_by(year, month) %>%
  summarise(roadkill_count = n(), .groups = "drop")

roadkill_with_climate <- roadkill_counts %>% #join counts and climate data
  left_join(climate_data, by = c("year", "month"))


get_season <- function(month) { #add season
  if (month %in% c(12, 1, 2)) {
    return("Winter")
  } else if (month %in% c(3, 4, 5)) {
    return("Spring")
  } else if (month %in% c(6, 7, 8)) {
    return("Summer")
  } else if (month %in% c(9, 10, 11)) {
    return("Autumn")
  } else {
    return(NA)
  }
}


roadkill_with_climate <- roadkill_with_climate %>%
  mutate(season = sapply(month, get_season))

roadkill_with_climate$season <- factor(roadkill_with_climate$season, 
                                       levels = c("Winter", "Spring", "Summer", "Autumn")) #convert to factor
```

## Seasonal Trends

Graph of seasonal trends

```{r}
seasonal_trends <- mammal_no_spike %>%
  group_by(season, year) %>%
  summarise(roadkill_count = n(), .groups = "drop")

ggplot(seasonal_trends, aes(x = year, y = roadkill_count, color = season)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Seasonal Trends in Mammalian Roadkill",
    x = "Year",
    y = "Number of Roadkill Reports",
    color = "Season"
  ) +
  theme_minimal()

```

```{r}
seasonal_data <- mammal_no_spike %>%
  group_by(year, season) %>%
  summarise(Count = n(), .groups = 'drop')

ggplot(seasonal_data, aes(x = season, y = Count)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Seasonal Variation in Roadkill Counts",
       x = "Season",
       y = "Roadkill Counts") +
  theme_minimal()
```

Determine which model to use

```{r}
#year only
model_year <- glm.nb(Count ~ year, data = seasonal_data)

#season only
model_season <- glm.nb(Count ~ season, data = seasonal_data)

#both year and season
model_both <- glm.nb(Count ~ year + season, data = seasonal_data)

#year and season with interaction
model_interaction <- glm.nb(Count ~ year * season, data = seasonal_data)

#compare models using AIC
AIC(model_year, model_season, model_both, model_interaction)

lrt_year_vs_season <- anova(model_year, model_season, test = "Chisq")
lrt_season_vs_both <- anova(model_season, model_both, test = "Chisq")
lrt_both_vs_interaction <- anova(model_both, model_interaction, test = "Chisq")


```

### Creating a table for report

```{r}
#set up dataframe for table
model_comparison <- data.frame(
  Model = c("Year", "Season", "Year + Season", "Year * Season"),
  AIC = c(AIC(model_year),
          AIC(model_season),
          AIC(model_both),
          AIC(model_interaction)),
  LogLik = c(logLik(model_year),
             logLik(model_season),
             logLik(model_both),
             logLik(model_interaction)),
  DF = c(attr(logLik(model_year), "df"),
         attr(logLik(model_season), "df"),
         attr(logLik(model_both), "df"),
         attr(logLik(model_interaction), "df")),
  LR_Stat = c(NA, 
              lrt_year_vs_season$`LR stat.`[2],
              lrt_season_vs_both$`LR stat.`[2],
              lrt_both_vs_interaction$`LR stat.`[2]),
  P_Value = c(NA, 
              lrt_year_vs_season$`Pr(Chi)`[2],
              lrt_season_vs_both$`Pr(Chi)`[2],
              lrt_both_vs_interaction$`Pr(Chi)`[2])
)



ft <- model_comparison %>% #use flextable to form table
  mutate(P_Value = ifelse(P_Value < 0.01, "<0.01", as.character(round(P_Value, 2)))) %>%
  flextable() %>%
  colformat_double(j = c(2:6), digits = 2) %>% 
  align(align = 'center', part = 'header') %>% 
  align(align = 'center', part = 'body') %>%
  bold(~P_Value == "<0.01", j = "P_Value")


ft

```

Now I've picked the best model, i get a summary of it which tells me which seasons are significant

```{r}
summary(model_season)
```

## Spatial Trend

I decided to make an interactive map of the roadkill reports per region in the uk.

```{r}
uk_regions <- st_read("ITL2_JAN_2025_UK_BFE.shp") #upload shapefile


roadkill_sf <- mammal_no_spike %>% #change crs type
  st_as_sf(coords = c("long", "lat"), crs = 4326)  


uk_regions <- st_transform(uk_regions, crs = 4326) #match crs type


roadkill_with_regions <- st_join(roadkill_sf, uk_regions, join = st_intersects) #join regions and roadkill

print(names(uk_regions))

spatial_trends <- roadkill_with_regions %>%
  group_by(ITL225NM) %>%  
  summarise(roadkill_count = n(), .groups = "drop")

spatial_trends_map <- uk_regions %>%
  mutate(NAME = as.character(ITL225NM)) %>% 
  left_join(as.data.frame(spatial_trends), by = "ITL225NM")

tmap_mode("view") #interactive map 
m1 <- tm_shape(spatial_trends_map) +
  tm_polygons("roadkill_count", 
              title = "Roadkill Reports", 
              palette = "YlOrRd", 
              border.alpha = 0.5,
              id = "ITL225NM")

m1
```

Static map to include in report

```{r}
tmap_mode("plot") #static map

m2 <- tm_shape(spatial_trends_map) +
  tm_polygons("roadkill_count", 
              title = "Roadkill Reports", 
              palette = "YlOrRd", 
              border.alpha = 0.5,
              id = "ITL225NM") +
  tm_layout(
    legend.position = c("left", "top"),  
    legend.text.size = 0.5,                 
    legend.title.size = 0.8                 
  )

m2
```

I wanted to look at seasonal trends but there were too many reports for the maps to show anything

```{r}

ggplot(roadkill_sf) +
  geom_sf(aes(colour = season), size = 1) +
  facet_wrap(~season) +
  labs(title = "Distribution of Reports by Season",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

Again, even split into month there is not visually much difference

```{r}
ggplot(roadkill_sf) +
  geom_sf(aes(colour = month), size = 1) +
  facet_wrap(~month) +
  labs(title = "Distribution of Reports by Season",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

Test normality

```{r}
hist(spatial_trends_map$roadkill_count, xlab = "Counts", breaks = 10) #histogram of data

qqnorm(spatial_trends_map$roadkill_count) 
qqline(spatial_trends_map$roadkill_count, col = "red")
```

Check for overdispersion

```{r}

mean_count <- mean(spatial_trends_map$roadkill_count)
var_count <- var(spatial_trends_map$roadkill_count)
print(mean_count)
print(var_count)
#mean is a lot smaller than variance so negative binomial distribvution is likely better but i will calculate the dispersion statistic for the poisson model
```

```{r}
poisson_model <- glm(roadkill_count ~ NAME, family = poisson(link = "log"), data = spatial_trends_map) #glm using poisson family

dispersion <- sum(residuals(poisson_model, type = "pearson")^2) / df.residual(poisson_model)
print(dispersion) #calculate overdispersion

plot(poisson_model)
```


```{r}
region_data$region_group <- ifelse(region_data$NAME %in% c( #group regions into "North" etc
  "Tees Valley", 
  "Northumberland, Durham and Tyne & Wear", 
  "Cumbria", 
  "Greater Manchester", 
  "Lancashire", 
  "Cheshire", 
  "Merseyside", 
  "North Yorkshire", 
  "South Yorkshire", 
  "West Yorkshire", 
  "East Yorkshire and Northern Lincolnshire"
), "North",
ifelse(region_data$NAME %in% c(
  "Derbyshire and Nottinghamshire", 
  "Leicestershire, Rutland and Northamptonshire", 
  "Lincolnshire", 
  "Herefordshire, Worcestershire and Warwickshire", 
  "Shropshire and Staffordshire", 
  "West Midlands", 
  "Gloucestershire and Wiltshire"
), "Midlands",
ifelse(region_data$NAME %in% c(
  "Bedfordshire and Hertfordshire", 
  "Essex", 
  "Cambridgeshire and Peterborough", 
  "Norfolk", 
  "Suffolk", 
  "Berkshire, Buckinghamshire and Oxfordshire", 
  "Surrey, East and West Sussex", 
  "Hampshire and Isle of Wight", 
  "Kent", 
  "Cornwall and Isles of Scilly", 
  "Devon", 
  "West of England", 
  "North Somerset, Somerset and Dorset", 
  "Outer London - East and North East", 
  "Outer London - South", 
  "Outer London - West and North West", 
  "Inner London - East", 
  "Inner London - West"
), "South",
ifelse(region_data$NAME %in% c(
  "North Wales", 
  "Mid and South West Wales", 
  "South East Wales"
), "Wales",
ifelse(region_data$NAME %in% c(
  "Eastern Scotland", 
  "East Central Scotland", 
  "Highlands and Islands", 
  "West Central Scotland", 
  "North Eastern Scotland", 
  "Southern Scotland"
), "Scotland",
 ifelse(region_data$NAME %in% c("Northern Ireland"), "Northern Ireland", NA))))))

region_data$region_group <- as.factor(region_data$region_group) #convert to factor

table(region_data$region_group) #group sizes

```

```{r}
nb_model <- glm.nb(roadkill_count ~ region_group, data = region_data) #moel for roadkill vs region

summary(nb_model)

par(mfrow = c(2,2))
plot(nb_model) #diagnostic plots
```


```{r}
pairwise_results <- emmeans(nb_model, pairwise ~ region_group, type = "response") #pairwise comparisons between region groups

summary(pairwise_results)

```

```{r}
emmeans_table <- as.data.frame(pairwise_results$emmeans) %>% #layout for table for report
  mutate(
    response = round(response, 2),          
    SE = round(SE, 2),                     
    asymp.LCL = round(asymp.LCL, 2),       
    asymp.UCL = round(asymp.UCL, 2)        
  ) %>%
  rename(
    `Region Group` = region_group,
    `Predicted Count` = response,
    `Standard Error` = SE,
    `95% CI (Lower)` = asymp.LCL,
    `95% CI (Upper)` = asymp.UCL
  )


et <- emmeans_table %>% #flextable
  flextable() %>%
  colformat_double(j = c("Predicted Count", "Standard Error", "95% CI (Lower)", "95% CI (Upper)"), digits = 2) %>% 
  align(align = 'center', part = 'header') %>% 
  align(align = 'center', part = 'body') %>%
  autofit()

et
```

## Role of temp and rainfall

```{r}

#try linear regression model
model <- lm(roadkill_count ~ mean_temperature + rainfall, data = roadkill_with_climate)


summary(model)

par(mfrow = c(2, 2))
plot(model) #check diagnostic plots
```
Data definitely not normal so use a glm instead - decide between poisson and negative binomial
```{r}
glm_poisson_climate <- glm(roadkill_count ~ mean_temperature + rainfall, family = poisson, data = roadkill_with_climate)

summary(glm_poisson_climate)

climate_dispersion <- sum(residuals(glm_poisson_climate, type = "pearson")^2) / df.residual(glm_poisson_climate)

climate_dispersion
```

Because it is over-dispersed, I will use a negative binomial family instead

```{r}
glm_nb_climate <- glm.nb(roadkill_count ~ mean_temperature + rainfall, data = roadkill_with_climate)

summary(glm_nb_climate)

par(mfrow = c(2, 2))
plot(glm_nb_climate)
```

```{r}
#plot temperature - scatterplot with regression line
temp_rk <- ggplot(roadkill_with_climate, aes(x = mean_temperature, y = roadkill_count)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "glm.nb", se = TRUE) +  
  labs(title = "Effect of Temperature on Roadkill Counts",
       x = "Temperature (°C)",
       y = "Roadkill Count")

temp_rk
```

```{r}
#same for rain
rain_rk <- ggplot(roadkill_with_climate, aes(x = rainfall, y = roadkill_count)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "glm.nb", se = TRUE) +  
  labs(title = "Effect of Rainfall on Roadkill Counts",
       x = "Rainfall (mm)",
       y = "Roadkill Count")

rain_rk
```
